\chapter*{Acknowledgements}\label{cha:ack}

I would like to extend my heartfelt thanks to the members of the jury, and especially to the reviewers, Smita Krishnaswamy and Gabriel Peyré, for their time and invaluable feedback.

My deepest gratitude goes to my advisors. First and foremost, to my PhD supervisor, Aurélien Garivier, for his invaluable guidance and unwavering support. I am deeply thankful to Titouan Vayer for the countless discussions, insightful feedback, and productive collaborations. I am deeply grateful to my close collaborators, Rémi Flamary, Nicolas Courty, and Cédric Vincent-Cuaz, for their invaluable scientific exchanges, advice, and stimulating discussions, as well as for their productive collaborations. I am very grateful for the trust they have placed in me.

My sincere thanks also go to Franck Picard, Julien Chiquet, and Thibault Espinasse for engaging in meaningful projects and discussions during the early stages of my PhD. That initial project we undertook together truly set me on this exciting path in dimensionality reduction methods, providing both direction and inspiration for my future research.

I extend my gratitude to all my colleagues at UMPA and LBMC, especially the team members of Olivier Gandrillon, for their collegiality and support.

I would like to thank Randall Balestriero for his confidence in me as we set up his lab at Brown. It has been a pleasure working with him, and I am grateful for the opportunity to continue our collaboration on future projects.

I am thankful to everyone who contributed to my growth through advice, engaging scientific discussions, or various forms of support: Rémi Gribonval, Bertrand Michel, Jean Feydy, Alain Durmus, Vivien Cabannes, Dana Pe'er, Guillaume Huguet, Brandon Amos, Emile Mathieu, Romain Lopez, and many others.

\newpage