\section*{Notations}

\begin{itemize}
    \item $\integ{n}$: set of integers $\{1,\dots,n\}$.
    \item $\mathbf{1}_n$: vector in $\R^n$ with all entries equal to $1$. Its size may be inferred from the context when omitted.
    \item $\mI_n$: identity matrix of size $n \times n$.
    \item $\exp(\cdot)$, $\log(\cdot)$: applied element-wise when used with vectors/matrices.
    \item $\mathcal{S}^n$: space of $n \times n$ symmetric matrices.
    \item $v_i$ or $[\vv]_i$: the $i$-th entry of a vector $\vv$. Similarly, $M_{ij}$ or $[\mM]_{ij}$: the $(i,j)$ entry of a matrix $\mM$.
    \item $\mathbf{P}_{i:}$ or $[\mathbf{P}]_{i:}$: the $i$-th row of a matrix $\Pb$.
    \item $|\mM|$: determinant of a square matrix $\mM$.
    \item $\odot$ and $\oslash$: element-wise multiplication and division, respectively, between vectors/matrices.
    \item $\langle \cdot, \cdot \rangle$: standard Euclidean inner product for vectors/matrices.
    \item $\bm{\alpha} \oplus \bm{\beta}$: for $\bm{\alpha}, \bm{\beta} \in \R^n$, this denotes the matrix in $\R^{n \times n}$ with $(i,j)$ entry given by $\alpha_i + \beta_j$.
    \item $\bm{p}^{\odot \alpha}$ or $\mP^{\odot \alpha}$: element-wise exponentiation, where $[\bm{p}^{\odot \alpha}]_i = p_i^\alpha$ and $[\mP^{\odot \alpha}]_{ij} = P_{ij}^\alpha$.
    \item $[\mP]_+$: the element-wise positive part of a matrix $\mP$, defined as $[\mP]_+ = \max(0, P_{ij})$.
    \item $\operatorname{H}(\p)$: entropy of $\p \in \mathbb{R}^{n}_+$, defined as $\operatorname{H}(\p) = -\sum_{i} p_i (\log(p_i) - 1) = -\langle \pb, \log \pb - \bm{1} \rangle$, with the convention $0 \log 0 = 0$.
    \item $L_2(x,y) \coloneqq \frac{1}{2} |x - y|^2$: quadratic loss.
    \item $L_{\mathrm{KL}}(x,y) \coloneqq x (\log(x) - \log(y)) - x + y$: Kullback-Leibler divergence loss.
    \item The Kullback-Leibler divergence between vectors and matrices are respectively given by $\mathrm{KL}(\va | \vb) = \sum_{i} \mathrm{KL}(a_{i} | b_{i})$ and $\mathrm{KL}(\mA | \mB) = \sum_{ij} \mathrm{KL}(A_{ij} | B_{ij})$.
    \item $S_N$: set of permutations of $\integ{N}$.
    \item $P_N(\R^d)$: set of discrete probability measures composed of $N$ points in $\R^d$.
    \item $\Sigma_N$: the probability simplex of size $N$, defined as $\Sigma_N := \{\vh \in \R_+^N \mid \sum_i h_i = 1\}$.
    \item $\diag(\vx)$: diagonal matrix with the elements of vector $\vx \in \R^N$ on the diagonal.
    \item $\gU(\vh)$: set of discrete couplings with one marginal, defined as $\gU(\vh) = \left\{ \mT \in \R_+^{N \times n} \mid \mT \mathbf{1}_n = \vh \right\}$.
    \item $\gU(\vh, \overline{\vh})$: set of discrete couplings with two marginals, defined as $\gU(\vh, \overline{\vh}) = \left\{ \mT \in \R_+^{N \times n} \mid \mT \mathbf{1}_n = \vh, \mT^\top \mathbf{1}_N = \overline{\vh} \right\}$.
    \item $\Delta^{n}$: the probability simplex $\{ \bm{p} \in \R_+^n \mid \sum_i p_i = 1 \}$.
    \item $\Pi(\bm{a}, \bm{b})$: the transport polytope with marginals $(\bm{a}, \bm{b})$, defined as $\Pi(\bm{a}, \bm{b}) = \{ \mP \in \R_+^{N_S \times N_T} \mid \mP \mathbf{1} = \bm{a}, \mP^\top \mathbf{1} = \bm{b} \}$.
    \item $\Pi(\bm{a})$: the semi-relaxed transport polytope, defined as $\Pi(\bm{a}) = \{ \mP \in \R_+^{N_S \times N_T} \mid \mP \mathbf{1} = \bm{a} \}$.
    \item $\operatorname{Proj}^{D}_{\mathcal{E}}(\mK)$: projection of $\mK$ onto a set $\mathcal{E}$ with respect to a divergence $D$, defined as $\operatorname{Proj}^{D}_{\mathcal{E}}(\mK)  =  \argmin_{\mP \in\mathcal{E}} D(\mP | \mK)$.
\end{itemize}